---
title: Create a pivot table in PostgreSQL
author: Yen Chun Chen
date: '2024-01-27'
slug: Create-a-pivot-table-in-PostgreSQL
categories:
  - Postgresql
tags:
  - R Markdown
  - Postgresql
subtitle: ''
summary: 'Illustrate how to use a pivot table to report data more clearly.'
authors: []
lastmod: '2024-01-27T14:52:35+08:00'
featured: no
image:
  caption: 'Friends scenes from Reddit'
  focal_point: 'smart'
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
---

Pivots don't always make people crazy! It could sometimes give us a clear view of the mess data, let's see how it works!

### Setup database table

We will use a dataset with New York City parking violation records.

```{r setup}
# Creating a temporal database, only RSQLite has this feature so far
# Connect to the default database
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "")
knitr::opts_chunk$set(connection = "con")
```

```{r include=FALSE}
# Load dataset
parking_violation <- read.csv("C:/Users/User/Documents/R Scripts/DataCamp/data/NYC_Open_Data_Parking_Violations_2020.sample5k.csv", na.strings = "")
```

```{r include=FALSE}
# Replace column name dots with _
colnames(parking_violation) <- stringr::str_replace_all(
    colnames(parking_violation), "\\.", "\\_")

colnames(parking_violation) <- stringr::str_to_lower(
    colnames(parking_violation))
```

```{r}
# Glimpse dataset
dplyr::glimpse(parking_violation)
```


```{r}
# Create database tables
dbWriteTable(con, "parking_violation", parking_violation)
dbListTables(con)
```


### Selecting data

Suppose we want to get a better understanding of which agencies (`issuing_agency`) are responsible for different types of parking violations (`violation_code`).

```{sql}
-- How many type of agency: 11
SELECT issuing_agency, COUNT(*) AS N
FROM parking_violation
GROUP BY issuing_agency
ORDER BY N DESC
```

These four agencies will be our main focus:

-   `Police Department` (`P`)

-   `Department of Sanitation` (`S`)

-   `Parks Department` (`K`)

-   `Department of Transportation` (`V`)

### Without a pivot table

Here we will write a `SELECT` query that contains three columns:

- parking violation code

- issuing agency code

- total number of records with each combination


```{sql}
SELECT 
	violation_code, 
    issuing_agency, 
    -- Number of records with violation code/issuing agency
    COUNT(*) 
FROM 
	parking_violation 
WHERE 
	-- Restrict the results to the agencies of interest
	issuing_agency IN ('P', 'S', 'K', 'V') 
GROUP BY 
	-- Define GROUP BY columns to ensure correct pair count
	violation_code, issuing_agency
ORDER BY 
	violation_code, issuing_agency;
```

The result shows the `violation_code` and `issuing_agency` pairs are repeated throughout the outcomes, ending up with 93 rows!

### With a pivot table

To generate the results in a pivot table format, we will use the `FILTER` clause. It is now easier to present the parking violation by agency for all four with this modified data format.

```{sql}
SELECT 
	violation_code, 
    -- Define the "Police" column
	COUNT(issuing_agency) FILTER (WHERE issuing_agency = 'P') AS "Police",
    -- Define the "Sanitation" column
	COUNT(issuing_agency) FILTER (WHERE issuing_agency = 'S') AS "Sanitation",
    -- Define the "Parks" column
	COUNT(issuing_agency) FILTER (WHERE issuing_agency = 'K') AS "Parks",
    -- Define the "Transportation" column
	COUNT(issuing_agency) FILTER (WHERE issuing_agency = 'V') AS "Transportation"
FROM 
	parking_violation 
GROUP BY 
	violation_code
ORDER BY 
	violation_code
```

When the results are shown in a pivot table, by just looking at the rows, it is possible to compare the number of violations for each agency.

```{r}
# Disconnect database
dbDisconnect(con)
```

*Notice: This content was modified from an exercise on DataCamp.*
